---
title: "Introduction to R"
author: "Matthew Thomas"
output: 
  beamer_presentation: 
    pandoc_args: "--highlight=style/my.theme"
    includes:
      in_header: style/twocol.tex
    theme: metropolis
---

```{r include=FALSE}
library(tidyverse)
set.seed(100)
knitr::opts_chunk$set(comment = NA)
# Adjust summary function so that results fit on the page
summary2 <- summary
summary <- function(object, ...){
  return(summary2(object, ...)$coef)
}

# Change ggplot2 theme
theme_set(theme_bw())
```

## Why R?

* R is free, open source, and incredibly popular
* There is a large (and welcoming) community of R programmers online who can help troubleshoot code and answer questions
* The language is incredibly well (and consistently) documented
* There are thousands of packages which implement statistical estimators and other use cases.

# Defining variables/basic data types

## Vectors and Assignment
The function `c()` takes vectors and creates a new longer vector. The assignment operator `<-` is a shortcut for the `assign()` function.
```{r}
x <- c(1,2,4,6,10:13)
assign("y",c(1,2,4,6,10:13))
```
```{r}
x
y
```

## Operators
```{r}
x/y # Operators on vectors apply element-wise
(1:2) * (1:8)  # Vectors will repeat if necessary
(1:6) > (6:1)  # Logical operators: <,<=,==,>=,>,!=
!(1:6) > (6:1) # Reverse logic with !
```

## Matrices
A matrix is a vector with a dimension attribute. Matrices are filled column by column unless specified.
```{r}
(mat <- matrix(data = x, ncol = 2))
```

## Sub-setting Matrices
You can subset a matrix using row,col indexing.
```{r}
mat[1,]  # First row of matrix
mat[,2]  # Second column of matrix
mat[1,2] # Second element of first row
```

## Warning about One Dimensional Matrices  
An nx1 matrix and a vector are not the same thing. For example, a nx1 matrix will not replicate if necessary.

```{r error=TRUE}
matrix(1:2) * matrix(1:8)
```

## Defining Functions
Functions are objects in R that can be applied to other objects. `c()`, `mean()`, and `sum()` are examples of built-in functions. You can also write your own functions.

```{r}
sumsq <- function(var){
  return(sum(var^2))
}
```

## Calling Functions
These functions can be called just as any built-in function.
```{r}
sumsq(c(1,2))
```

The convenience operator `%>%` passes the preceding object to the first argument of any function.
```{r}
c(1,2) %>% sumsq()
```

## Lists
Lists can contain any object types.
```{r}
z <- list( "y" = y,
           "istwo" = y^2 == y*2,
           "p" = runif(8)*(1:4)/y^2 )
```
You can reference items from a list using brackets or dollar sign
```{r}
z["y"]  # Returns a single element list
z$istwo # Returns a vector
```

# Dealing with data frames

## Creating a data frame
You can make a data frame using vectors or a list. Data frames are special lists with elements of the same length.
```{r}
(df1 <- data.frame(z))
```

## Adding to data frames
You can reference and add to a data frame just as you can with any other list. However, data frames will repeat elements if necessary to enforce the length requirement.
```{r}
df1$prod <- LETTERS[1:4]
head(df1)
```
## Matrix-like properties of data frames

Due to the length requirement, data frames have limited matrix like properties. You can index a data frame just like a matrix.
```{r}
df1[1,] # First row of data frame
```
You can even apply most operators to **numeric** data frames. Linear algebra operators do not work on data frames.
```{r}
df1[1,1:3]+1 # Have to exclude prod
```

## Manipulating data frames
You can manipulate data using the traditional list interface
```{r}
df1$ly <- log(df1$y)
```
The `tidyverse` package has introduced another way to do this using the `mutate()` function
```{r}
df1 <- df1 %>% mutate(ly2 = log(y))
head(df1,4)
```

# Regression

## Running a regression
If you just want to run a regression in R, often do not need to manipulate data. Regressions in R allow you to adjust variables using ``formulas''. Suppose we want to estimate the following model:
$$
\log(y) = \beta_0 + \beta_1 \log(p) + \beta_2 prodB + \beta_3 prodC + \beta_4 prodD
$$
\small
```{r}
lm(log(y) ~ log(p) + prod, data = df1) %>% summary()
```
\normalsize

## Interaction terms
You can add interaction terms by using a `:` between two variable names.
\small
```{r}
lm(log(y) ~ log(p) + log(p):prod, data = df1) %>% summary()
```
\normalsize

## Removing the constant
You can suppress the constant by adding `-1` to the formula. Note that it automatically adds the dummy for product A back into the regression.
\small
```{r}
lm(log(y) ~ log(p) + prod - 1, data = df1) %>% summary()
```
\normalsize

## Polynomials
R does not allow arbitrary binary operators inside of an equation.
\small
```{r}
lm(log(y) ~ p + p^2, data = df1) %>% summary()
```
\normalsize
To run a polynomial fit, you need to use the `poly` function
\small
```{r}
lm(log(y) ~ poly(p,2), data = df1) %>% summary()
```
\normalsize

## Overriding
But what if you just want the square term? For that, you need to override using the inhibit function, `I()`.
\small
```{r}
lm(log(y) ~ p^2, data = df1) %>% summary()
lm(log(y) ~ I(p^2), data = df1) %>% summary()
```
\normalsize

# Visualization

## Builtin graphics
There are several basic builtin plot commands builtin to R.

\begincols
  \begincol{.48\textwidth}
```{r}
plot(df1$y ~ df1$p, cex=2)
```
  \endcol
\begincol{.48\textwidth}
```{r}
hist(df1$p, breaks = 8)
```
  \endcol
\endcols

They are not very pretty, but they are very easy to use.

## ggplot2 graphics

```{r fig.height = 4.5}
ggplot(data = df1, aes(x=p, y=y, col=prod)) + 
  geom_point(size=2) +                      
  geom_smooth(method="lm", col="blue", size=1) + 
  coord_cartesian(xlim=c(0,0.3), ylim=c(0,13)) + 
  labs(title="Demand", y="Quantity", x="Price")
```


# Appendix

## Converting a data frame to a matrix
Because a matrix can contain categorical variables and strings, it is not always possible to directly convert a data frame to a matrix. An all numeric data frame can be converted by simply using `as.matrix()`
```{r}
dfa <- data.frame(a=1:5,b=77:81,c=log(22:18))
dfb <- data.frame(a=letters[1:5],b=77:81,c=log(22:18))
```
\begincols
  \begincol{.48\textwidth}
```{r}
as.matrix(dfa)
```
  \endcol
\begincol{.48\textwidth}
```{r}
as.matrix(dfb)
```
  \endcol
\endcols

## Converting a data frame to a matrix
In order to properly convert a data frame with strings or factors into a numeric matrix, we need to use `model.matrix()`. This is what R uses when it runs regressions.
```{r}
model.matrix(~a+b+c-1,dfb)
```

## Linear Algebra and apply
The `apply()` function applies some function across rows `(MARGIN=1)` or columns `(MARGIN=2)` of a matrix.
```{r}
apply(X=mat, MARGIN=1, FUN=sumsq)
```

The operators `%*%` and `%^%` do matrix multiplication and exponentiation. The function `t()` transposes. If you can accomplish a task with linear algebra, it is generally faster than `apply()`. 
```{r}
c(mat^2 %*% c(1,1))
```
for example is more than twice as fast for a large matrix.

